Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex
an actual disco now

It was once a major undertaking to collect data sufficent to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
Bypassing this bottleneck will require new data structures appropriate for genomic data.

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them.
Above, we showed that branch and site statistics matched well in simulated data,
but these simulations made some simplifying assumptions, including constant mutation rates.
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences:
by \tsinfer{} \citep{kelleher2019inferring}, which only aims to estimate tree topologies,
not well-calibrated node times,
(2) by GEVA \citep{albers2019dating}, which aims to calibrate the node times reported by \tsinfer,
and (3) by Relate \citep{speidel2019method}, an independently developed method.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.

@@

Thus far we have examined how statistics vary along the genome,
but a tree sequence allows us access to another dimension: time.
Next, we examine how the \emph{time} distribution of admixture signal
in the tree sequence inferred by \tsinfer/GEVA.
The branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
A branch with a large value is expected to contribute a large amount to the Site $f_4$,
which is calculated from genotyping data
and is used to infer historical connections between populations.
To visualize the time signature, we assigned the contribution of each branch
to its terminal node,
and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
We then normalised these to sum to 1 by dividing by the total Branch $f_4$ value.

@@

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$. For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the 
        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
        \label{fig:stats_performance}
    }
\end{figure}

@@

% this stuff is faster
The most direct effect of the methods presented here might be on efficiency of computation
with large genomic datasets --
the python library \tskit{} provides fast, rigorously tested implementations
of many standard population genetic statistics, with more on the way.
The logarithmic scaling of storage and computation time with sample size
could bring much more computation-intensive methods within reach on modern large genomic datasets
of humans, 
and helps make detailed investigation of whole populations in evolutionary simulations feasible.

% but the duality thing is maybe more long-term exciting
However, the focus of this paper was not on computation of standard population genetics statistics,
but on the relationship of these statistics to the underlying trees.
We have used essentially the same general framework
to define two additional classes of statistics -- Branch and Node statistics --
that well-defined relationships to the Site statistics computable from genome sequence.
% describing trees is cool
At least, this gives us descriptive tools to directly describe the genealogical trees
of tree sequences obtained by simulation.
% and we can do it with real data now
However, new methods to infer tree sequences allow us to apply these methods to real data.
% branch stats are less noisy!
Given the true tree sequence,
Branch statistics are less-noisy versions of the standard Site statistics,
so methods that use Branch statistics may show increased power and precision
over the corresponding genotype-only methods.
We have shown that this is true in simulated data,
but it is not yet known how much additional noise the process of tree sequence inference will add.
% and we get the time dimension
Furthermore, tree sequences add a \emph{time} dimension to genotype data,
which these descriptive methods open up many possibilities for utilizing and visualizing of.

\paragraph{Future work}
In this paper,
we have considered only so-called additive, single site statistics.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
such as linkage disequlibrium are planned for future work.
Haplotype-based statistics may require a different class of algorithms.

@@

\begin{figure}
    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
    \includegraphics[width=\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} \tsinfer,
        \textbf{(bottom left)} GEVA, and
        \textbf{(bottom right)} Relate.
        \label{fig:chr20_full}
    }
\end{figure}

>>>>

It was once a major undertaking to collect data sufficient to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
Bypassing this bottleneck will require new data structures appropriate for genomic data.

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them.
Above, we showed that branch and site statistics matched well in simulated data,
but these simulations made some simplifying assumptions, including constant mutation rates.
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences:
by \tsinfer{} \citep{kelleher2019inferring}, which only aims to estimate tree topologies,
not well-calibrated node times,
(2) by GEVA \citep{albers2019dating}, which aims to calibrate the node times reported by \tsinfer,
and (3) by Relate \citep{speidel2019method}, an independently developed method.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
Separate curves for each of the 26 sampling locations are shown in Supplemental Figure \ref{fig:chr20_full}.

@@

Thus far we have examined how statistics vary along the genome,
but a tree sequence allows us access to another dimension: time.
Next, we examine the \emph{time} distribution of admixture signal
in the tree sequence inferred by \tsinfer/GEVA.
The branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
A branch with a large value is expected to contribute a large amount to the Site $f_4$,
which is calculated from genotyping data
and is used to infer historical connections between populations.
To visualize the time signature, we assigned the contribution of each branch
to its terminal node,
and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
We then normalised these to sum to 1 by dividing by the total Branch $f_4$ value.

@@

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
        per generation, respectively.
        For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the 
        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
        \label{fig:stats_performance}
    }
\end{figure}

@@

% overview: general definition that is mostly functions of jAFS but includes others
In this paper, we have described a general framework for summarising genetic variation
and the underlying genealogies
that encompasses many standard population genetic statistics
and makes it easy to explore more.
Many of these statistics are functions of the joint allele frequency spectrum,
but the framework is more general,
and can be used to quantify associations between genotypes and phenotypes.
This generality was motivated in part by laziness --
adding new statistics is now almost as easy as working out the appropriate summary function --
but also makes it easy to computationally explore new classes of statistics.
The range of statistics available to describe variation in a single exchangeable sample
(i.e., a single population) are well-understood,
but there are much larger and less well-explored classes of statistics
describing genetic variation between many populations
or across geographical space.

% speed/efficiency: big data sets; large simulations
The most obvious application of these methods on current practice
is to improve efficiency of existing pipelines,
as tree sequences allow storage and processing of large genomic data sets
with orders of magnitude less space and time than most other methods.
All statistics described here (and more) are implemented
in the fast, rigorously tested python library \tskit{}
that also provides tools for working with tree sequences.
The largest efficiency gains are found for very large sample sizes
only likely to be available in real data sets for humans in the near future,
but the ability to analyze millions of whole genomes
is useful for understanding evolutionary simulations of any species
\citep[see][for a recent example]{galloway2019stickleback}.

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is this population genetics statistic summarising?
This can help when interpreting results, especially if reality may not fit an idealised model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to infer the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
essentially uses the signal from nearby patterns of variation
to interpret relationships at a given site,
thus tranforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed along two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
In this paper, we have made an exploratory foray in this direction,
by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
between several subsets of the 1000 Genomes dataset,
in several inferred tree sequences.
It is not yet clear whether these time profiles tell us more about history
or about biases of the the inference methods,
but either result would be useful.

% moving past a few summaries in discrete groups to a genome/time/space view of relatedness
The historical toolbox of population genetics
is still mostly composed of statistics designed for analysis of a handful of loci
in a small number of discrete, mostly separated populations.
Both our data and our understanding of the world have moved beyond this.
We hope that the tools we develop here will help make it possible
to visualize and analyze genetic variation and genealogical realtedness
along the genome, across space, and through time.
\plr{okay rein me in here}

\paragraph{Future work}
In this paper,
we have considered only so-called additive, single site statistics.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
such as linkage disequlibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} \tsinfer,
        \textbf{(bottom left)} GEVA, and
        \textbf{(bottom right)} Relate.
        \label{fig:chr20_full}
    }
\end{figure}
